---
title: "Instructor Guide: Data Wrangling"
author: "Foundations of Business Analytics"
format: html
---

## Opening framing

The goal today is not to show students lots of code. It is to show them how a few core `dplyr` tools help answer a business question.

The question for the lecture is:

**Which ASX industries appear strongest on return on invested capital (ROIC) in 2024?**

### What to say

* Today is about using code to answer a question, not code for its own sake.
* The key mental model is: **rows → columns → groups → answer**.
* We are going to move slowly.
* The point is not to memorise everything at once. The point is to understand what each step is doing to the data.

## Load packages

```{r}
library(tidyverse)
library(scales)
library(ggokabeito)
```

### What to say

* `tidyverse` gives us the core tools for the lecture.
* `scales` helps with readable labels later on.
* `ggokabeito` gives us a colour-blind-friendly palette, which matters when we present results.

## Load the data

```{r}
asx_200_2024 <- read_csv("../data/asx_200_2024.csv")
```

### What to say

* We are starting with firm-level data on large ASX companies.
* Before touching anything, we should look at what is actually in the dataset.

## Inspect the data

```{r}
slice_head(asx_200_2024, n = 10)
```

```{r}
glimpse(asx_200_2024)
```

### What to say

* `slice_head()` gives us a quick look at the first few rows.
* `glimpse()` shows us the variable names and data types.
* Right now we are just getting our bearings.
* The variables we care about for today are `ebit`, `invested_capital`, and `industry`.

## `filter()`

### What to say before coding

* `filter()` keeps rows that satisfy a condition.
* This is the rows part of the lecture.
* We will start simple, then build toward the version we actually need.

```{r}
positive_investment <- filter(asx_200_2024, invested_capital > 0)
head(positive_investment)
```

### What to say

* This keeps only firms where `invested_capital > 0` is true.
* The main idea is simple: if the condition is false, the row is dropped.

```{r}
missing_ebit <- filter(asx_200_2024, is.na(ebit))
head(missing_ebit)
```

### What to say

* `is.na(ebit)` means “show me the rows where EBIT is missing.”
* Real data often has missing values, so this is a very practical pattern to know.

```{r}
nonmissing_ebit <- filter(asx_200_2024, !is.na(ebit))
head(nonmissing_ebit)
```

### What to say

* The `!` symbol means “not.”
* So `!is.na(ebit)` means “keep rows where EBIT is available.”

```{r}
asx_clean <- filter(asx_200_2024, !is.na(ebit), invested_capital > 0)
```

### What to say

* We can combine conditions inside `filter()`.
* The comma here acts like AND.
* So we are keeping firms with non-missing EBIT **and** positive invested capital.
* This gives us a cleaner starting point for calculating ROIC.

## `mutate()`

### What to say before coding

* `mutate()` creates new columns or changes existing ones.
* This is the columns part of the lecture.

```{r}
asx_roic <- mutate(asx_clean, roic = ebit / invested_capital)
```

### What to say

* Here we create a new variable called `roic`.
* This is one of the most common uses of `mutate()`: build a new column from existing columns.

```{r}
slice_head(
  select(asx_roic, conml, industry, ebit, invested_capital, roic),
  n = 10
)
```

### What to say

* Pause here and make sure students can see that the new column now exists.
* Also stress that we are still at the firm level. We have added a column, not changed the unit of analysis.

```{r}
asx_roic_names <- mutate(asx_roic, conml_upper = str_to_upper(conml))
```

```{r}
slice_head(
  select(asx_roic_names, conml, conml_upper, roic),
  n = 10
)
```

### What to say

* `mutate()` is not only for arithmetic.
* It can also apply functions to an existing column.
* Here, `str_to_upper()` changes company names to uppercase.
* The broader point is that `mutate()` is a general tool for creating transformed variables.

## `|>`

### What to say before coding

* So far, we have gone one step at a time.
* That is often the right way to work when learning or checking your work.
* But sometimes the steps clearly belong together.
* That is where the pipe helps.

```{r}
asx_roic_pipe <- asx_200_2024 |>
  filter(!is.na(ebit), invested_capital > 0) |>
  mutate(roic = ebit / invested_capital)
```

```{r}
asx_roic_pipe |>
  select(conml, industry, ebit, invested_capital, roic) |>
  slice_head(n = 10)
```

### What to say

* The pipe takes the result from one step and passes it into the next step.
* This lets us write the workflow in the order the steps happen.
* It is not the only valid style, but it is often easier to read.

```{r}
slice_head(
  select(
    mutate(
      filter(asx_200_2024, !is.na(ebit), invested_capital > 0),
      roic = ebit / invested_capital
    ),
    conml, industry, ebit, invested_capital, roic
  ),
  n = 10
)
```

```{r}
asx_200_2024 |>
  filter(!is.na(ebit), invested_capital > 0) |>
  mutate(roic = ebit / invested_capital) |>
  select(conml, industry, ebit, invested_capital, roic) |>
  slice_head(n = 10)
```

### What to say

* Nested functions work, but they can be awkward because you have to read them from the inside out.
* The pipe lets you read the same logic step by step.
* Some students will prefer one style over the other, but this comparison usually makes the value of the pipe click.

## `group_by()` + `summarise()` + `arrange()`

### What to say before coding

* Now we move from firm-level data to industry-level summaries.
* This is the groups part of the lecture.
* This is also the point where the data starts answering the business question more directly.

```{r}
industry_roic <- asx_roic |>
  group_by(industry) |>
  summarise(
    avg_roic = mean(roic, na.rm = TRUE),
    n_firms = n(),
    .groups = "drop"
  ) |>
  arrange(desc(avg_roic))
```

```{r}
industry_roic
```

### What to say

* `group_by(industry)` divides the data into industry groups.
* `summarise()` then collapses each group into one row.
* `avg_roic` is the average ROIC within an industry.
* `n_firms = n()` tells us how many firms are contributing to that average.
* `arrange(desc(avg_roic))` ranks the industries from high to low.
* This is a key conceptual moment: we have changed the unit of analysis. We are no longer looking at firms; we are looking at industries.

```{r}
industry_roic_5plus <- industry_roic |>
  filter(n_firms >= 5)
```

```{r}
industry_roic_5plus |>
  slice_head(n = 10)
```

### What to say

* Some industries have very few firms, so their averages may be less informative.
* This step keeps industries with at least 5 firms.
* It is a simple example of analysts using judgment rather than treating every summary as equally reliable.

## `if_else()`

### What to say before coding

* `if_else()` is useful when we want two categories.
* This is a nice way to show that we can turn a numeric result into a simple label.

```{r}
industry_roic_5plus <- industry_roic_5plus |>
  mutate(
    roic_vs_median = if_else(
      avg_roic >= median(avg_roic, na.rm = TRUE),
      "Above median",
      "Below median"
    )
  )
```

```{r}
industry_roic_5plus |>
  select(industry, avg_roic, roic_vs_median) |>
  slice_head(n = 10)
```

### What to say

* The median gives us a data-driven cutoff.
* `if_else()` asks a yes/no question and returns one of two labels.

## `case_when()`

### What to say before coding

* `case_when()` is useful when we want more than two categories.
* Here we use quartiles so the cutoffs come from the data.

```{r}
roic_q25 <- quantile(industry_roic_5plus$avg_roic, 0.25, na.rm = TRUE)
roic_q50 <- quantile(industry_roic_5plus$avg_roic, 0.50, na.rm = TRUE)
roic_q75 <- quantile(industry_roic_5plus$avg_roic, 0.75, na.rm = TRUE)
```

```{r}
industry_roic_5plus |>
  mutate(
    performance_band = case_when(
      avg_roic >= roic_q75 ~ "Top quartile",
      avg_roic >= roic_q50 ~ "Upper-middle",
      avg_roic >= roic_q25 ~ "Lower-middle",
      TRUE ~ "Bottom quartile"
    )
  ) |>
  select(industry, avg_roic, performance_band) |>
  slice_head(n = 10)
```

### What to say

* `case_when()` checks several conditions in order.
* It is a flexible way to build multi-category labels from a continuous variable.

## `group_by()` with combinations

### What to say before coding

* `group_by()` can take more than one variable.
* To make that meaningful, we go back to the firm-level data.

```{r}
asx_roic_flag <- asx_roic |>
  mutate(
    roic_vs_median = if_else(
      roic >= median(roic, na.rm = TRUE),
      "Above median",
      "Below median"
    )
  )
```

```{r}
industry_roic_split <- asx_roic_flag |>
  group_by(industry, roic_vs_median) |>
  summarise(
    n_firms = n(),
    .groups = "drop"
  )
```

```{r}
industry_roic_split
```

### What to say

* Now the groups are combinations of industry and ROIC category.
* So instead of one summary per industry, we get subgroup counts within industries.

## Summary plot

### What to say before coding

* At this point the wrangling is doing the real work.
* The plot is there to communicate the result clearly, not to be the main event.

```{r}
top10_industries <- industry_roic_5plus |>
  slice_head(n = 10) |>
  mutate(industry = fct_reorder(industry, avg_roic))
```

```{r}
top10_industries |>
  ggplot(aes(x = industry, y = avg_roic, fill = industry)) +
  geom_col(show.legend = FALSE) +
  scale_fill_okabe_ito() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Average ROIC",
    title = "Top ASX Industries by Average Return on Invested Capital",
    subtitle = "Industries with at least 5 firms, 2024"
  ) +
  theme_minimal(base_size = 13)
```

### What to say

* The plot is not the main lesson; the wrangling is.
* But a clear plot is often the best way to present the final answer.
* This is a good place to ask: what do you notice, and how would you explain this to a manager?

## Closing wrap-up

### What to say

* The main point today is that a small number of `dplyr` functions can take us a long way.
* The core workflow was: **rows → columns → groups → answer**.
* `filter()` changed which rows we kept.
* `mutate()` created new variables.
* `group_by()` and `summarise()` moved us from firm-level data to industry-level summaries.
* `arrange()` helped us rank the result.
* The broader habit we want is not just writing code. It is using code to turn raw data into an answer someone can actually use.
